---
title: "p8105_hw3_xz3173"
author: "Xue Zhang"
date: "2023-10-05"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
```


# Problem 1

This problem uses the Instacart data. DO NOT include this dataset in your local data directory; instead, load the data from the p8105.datasets using:

```{r}
library(p8105.datasets)
data("instacart")
```


The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illustrative examples of observations. Then, do or answer the following (commenting on the results of each):

```{r}
head(instacart)
```



* How many aisles are there, and which aisles are the most items ordered from?

```{r}
# How many number of aisles are there?
num_aisles = length(unique(pull(instacart, aisle)))
num_aisles

# Which aisles are the most items ordered from?
aisle_orders = instacart |>
  group_by(aisle) |>
  summarise(order_count = n()) |>
  arrange(-order_count)

head(aisle_orders)
```

`Fresh vegetables` are the most items ordered from.

* Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r}
# filtering aisles with more than 10000 items ordered
instacart |>
  group_by(aisle) |>
  summarise(order_count = n()) |>
  filter(order_count > 10000) |>
  arrange(order_count) |>

# plotting
  ggplot(aes(x = reorder(aisle, order_count), y = order_count)) +
  geom_bar(stat = "identity") +
  coord_flip() + # to make it horizontal for better readability
  labs(title = "Number of Items Ordered in Each Aisle",
       x = "Aisle",
       y = "Number of Items Ordered") +
  theme_minimal()
```

```{r}
instacart |>
  group_by(aisle) |>
  summarise(order_count = n()) |>
  filter(order_count > 10000) |>
  arrange(order_count) |>
  ggplot(aes(x = reorder(aisle, order_count), y = order_count)) +
  geom_point() +
  coord_flip() +
  labs(title = "Number of Items Ordered in Each Aisle",
       x = "Aisle",
       y = "Number of Items Ordered") +
  theme_minimal()
```



* Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
# defining aisles of interest
aisles_of_interest = c("baking ingredients", "dog food care", "packaged vegetables fruits")

# create the table
popular_items = instacart |>
  filter(aisle %in% aisles_of_interest) |>
  group_by(aisle, product_name) |>
  summarise(order_count = n()) |>
  arrange(aisle, desc(order_count)) |>
  
  # keep the three most popular items in each of the aisles
  group_by(aisle) |>
  filter(min_rank(desc(order_count)) <= 3)

popular_items
```



* Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).


```{r}
# filter the data for the two products and calculate the mean order hour
product_data = instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarise(mean_hour = mean(order_hour_of_day, na.rm = TRUE)) |>
  mutate(day_name = factor(order_dow,
                           levels = 0:6,
                           labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) |>
  select(-order_dow)

# reshape the data for a 2 by 7 table format
table_data = product_data |>
  group_by(product_name) |>
  pivot_wider(
    names_from = day_name,
    values_from = mean_hour)

table_data
```


The `instacart` dataset has `r nrow(instacart)` observations and `r ncol(instacart)` variables. There are `r num_aisles` aisles are there, and `fresh vegetables` aisles are the most items ordered from. 



# Problem 2

Import data

```{r}
data("brfss_smart2010") 
```

First, do some data cleaning:

```{r}
brfss_smart2010 = brfss_smart2010 |> 
  
  # format the data to use appropriate variable names
  janitor::clean_names() |>
  
  # focus on the “Overall Health” topic
  filter(topic == "Overall Health") |>
  
  # include only responses from “Excellent” to “Poor”
  filter(response %in% c("Excellent", "Poor")) |>
  
  # organize responses as a factor taking levels ordered from “Poor” to “Excellent”
  mutate(response = factor(response, 
                           levels = c("Poor", "Excellent"),
                           ordered = TRUE))
```

* In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r}
brfss_smart2010 |>
  group_by(year, locationabbr) |>
  summarise(n_locations = n_distinct(locationdesc)) |>
  filter((year == 2002 | year == 2010) & n_locations >= 7)
```

* Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r}
excellent_df = brfss_smart2010 |>
  filter(response == "Excellent") |>
  group_by(year, locationabbr) |>
  summarise(average_value = mean(data_value, na.rm = TRUE))

ggplot(excellent_df, aes(x = year, y = average_value, group = locationabbr, color = locationabbr)) +
  geom_line() +
  labs(title = "Average Value Over Time for 'Excellent' Response", 
       x = "Year", 
       y = "Average Value") +
  theme_minimal()

```

* Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

```{r}
ny_df = brfss_smart2010 |>
  filter(locationabbr == "NY" & (year == 2006 | year == 2010))

ggplot(ny_df, aes(x = response, y = data_value)) +
  geom_boxplot() +
  facet_wrap(~year, ncol = 2) +
  labs(title = "Distribution of Data Value in NY for 2006 and 2010", 
       x = "Response", 
       y = "Data Value") +
  theme_minimal()
```

# Problem 3

```{r}
nhanes_covar =
  # import dataset
  read_csv("data/nhanes_covar.csv", skip = 4) |>
  
  # clean names
  janitor::clean_names() |>
  
  # ensure variables are appropriate encoded
  
  mutate(
    sex = 
      case_match(
        sex,
        1 ~ "Male",
        2 ~ "Female"),
    
    education =
      case_match(
        education,
        1 ~ "Less than high school",
        2 ~ "High school equivalent",
        3 ~ "More than high school")) |>
  
  # exclude participants less than 21 years of age
  filter(age == "." | age >= 21) 
  
  # exclude those with missing demographic data

```





```{r}
nhanes_accel =
  read_csv("data/nhanes_accel.csv") |>
  janitor::clean_names()
```

Accelerometers have become an appealing alternative to self-report techniques for studying physical activity in observational studies and clinical trials, largely because of their relative objectivity. During observation periods, the devices can measure MIMS in a short period; one-minute intervals are common. Because accelerometers can be worn comfortably and unobtrusively, they produce around-the-clock observations.

This problem uses accelerometer data collected on 250 participants in the NHANES study. The participants’ demographic data can be downloaded here, and their accelerometer data can be downloaded here. Variables *MIMS are the MIMS values for each minute of a 24-hour day starting at midnight.

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.


