---
title: "p8105_hw3_xz3173"
author: "Xue Zhang"
date: "2023-10-05"
output: github_document
---

```{r}
library(tidyverse)
library(ggridges)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

Load the datasets.

```{r}
library(p8105.datasets)
data("instacart")
```


The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illustrative examples of observations. Then, do or answer the following (commenting on the results of each):

```{r}
head(instacart)
```


* How many aisles are there, and which aisles are the most items ordered from?

```{r}
# How many number of aisles are there?
num_aisles = length(unique(pull(instacart, aisle)))
num_aisles

# Which aisles are the most items ordered from?
aisle_orders = instacart |>
  group_by(aisle) |>
  summarise(order_count = n()) |>
  arrange(-order_count)

head(aisle_orders)
```

`Fresh vegetables` are the most items ordered from.

* Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.



```{r}
# filtering aisles with more than 10000 items ordered
instacart |>
  group_by(aisle) |>
  summarise(order_count = n()) |>
  filter(order_count > 10000) |>
  arrange(order_count) |>

# plotting
  ggplot(aes(x = reorder(aisle, order_count), y = order_count)) +
  geom_bar(stat = "identity") +
  coord_flip() + # to make it horizontal for better readability
  labs(title = "Number of Items Ordered in Each Aisle",
       x = "Aisle",
       y = "Number of Items Ordered",
       caption = "Data from the instacart package") 

```
  
* Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
# defining aisles of interest
aisles_of_interest = c("baking ingredients", "dog food care", "packaged vegetables fruits")

# create the table
popular_items = instacart |>
  filter(aisle %in% aisles_of_interest) |>
  group_by(aisle, product_name) |>
  summarise(order_count = n()) |>
  arrange(aisle, desc(order_count)) |>
  
  # keep the three most popular items in each of the aisles
  group_by(aisle) |>
  filter(min_rank(desc(order_count)) <= 3)

popular_items
```



* Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).


```{r}
# filter the data for the two products and calculate the mean order hour
product_data = instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarise(mean_hour = mean(order_hour_of_day, na.rm = TRUE)) |>
  mutate(day_name = factor(order_dow,
                           levels = 0:6,
                           labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) |>
  select(-order_dow)

# reshape the data for a 2 by 7 table format
table_data = product_data |>
  group_by(product_name) |>
  pivot_wider(
    names_from = day_name,
    values_from = mean_hour) 

table_data
```


The `instacart` dataset has `r nrow(instacart)` observations and `r ncol(instacart)` variables. There are `r num_aisles` aisles are there, and `fresh vegetables` aisles are the most items ordered from. 



# Problem 2

Import data

```{r}
data("brfss_smart2010") 
```

First, do some data cleaning:

```{r}
brfss_smart2010 = brfss_smart2010 |> 
  
  # format the data to use appropriate variable names
  janitor::clean_names() |>
  
  # focus on the “Overall Health” topic
  filter(topic == "Overall Health") |>
  
  # include only responses from “Excellent” to “Poor”
  filter(response %in% c("Excellent", "Very good", "Good", "Fair", "Poor")) |>
  
  # organize responses as a factor taking levels ordered from “Poor” to “Excellent”
  mutate(response = factor(response, 
                           levels = c("Poor", "Fair", "Good", "Very good", "Excellent"),
                           ordered = TRUE))
```

* In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r}
states_with_7_or_more_locations_in_2002 = brfss_smart2010 |>
  filter(year == 2002) |>
  group_by(year, locationabbr) |>
  summarize(n_locations = n_distinct(locationdesc)) |>
  filter(n_locations >= 7) 

print(states_with_7_or_more_locations_in_2002)
```



```{r}
states_with_7_or_more_locations_in_2010 = brfss_smart2010 |>
  filter(year == 2010) |>
  group_by(year, locationabbr) |>
  summarize(n_locations = n_distinct(locationdesc)) |>
  filter(n_locations >= 7) 

print(states_with_7_or_more_locations_in_2010)
```

**CT, FL, MA, NC, NJ, PA are the states were observed at 7 or more locations in 2002. CA, CO, FL, MA, MD, NC, NE, NJ, NY, OH, PA, SC, TX, WA are the states with 7 or more locations in 2010. In 2002, there are 6 states had observations from 7 or more locations, which indicates the health assessments was more concentrated in these states in 2002. Comparing with that, there are 14 states had observations from 7 or more locations, which indicates the health assessment had a expansion in coverage region. CT is the only one state that list in 2002, butt not on 2010. This might cause by various reasons, such as funding loss or reduce, possible data collection loss or reduce, health priorities/focus transfer to other areas.**


* Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r}
excellent_df = brfss_smart2010 |>
  filter(response == "Excellent") |>
  group_by(year, locationabbr) |>
  summarize(average_value = mean(data_value, na.rm = TRUE)) 

print(excellent_df)
```

```{r}
ggplot(excellent_df, aes(x = year, y = average_value, group = locationabbr, color = locationabbr)) +
  geom_line(alpha = 0.8) +
  labs(title = "Average Value Over Time for 'Excellent' Response", 
       x = "Year", 
       y = "Average Value",
       caption = "Data from the brfss_smart2010 package") +
  theme(legend.position = "none") # to hide the legend since there will be many states
```


**The spaghetti plot displays the trend of "Excellent" health responses over the years for each state. (Since there is so many states, the legend is hid in the plot.) States with increasing trends show improving perceptions of health, while decreasing trends might indicate worsening perceptions. From the plot, we could see most states have volatile pattern with increasing and decreasing trends during 2002 to 2010. This indicates that variable factors influencing these perceptions, as well as differ from each states. The various factors might be different social determinants of health, such as economic, social, or environmental. For example, the improvements in access to healthcare services might shows the increasing trends. Unemployment might be one of the factor that impact the decreasing tends.**

* Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.



```{r}
ny_df = brfss_smart2010 |>
  filter(locationabbr == "NY"  & (year == 2006 | year == 2010)) 

ggplot(ny_df, aes(x = response, y = data_value)) +
  geom_boxplot() +
  facet_grid(. ~ year) +
  labs(title = "Distribution of Data Value in NY for 2006 and 2010", 
       x = "Response", 
       y = "Data Value",
       caption = "Data from the brfss_smart2010 package")
  
```

**The two-panel plot shows the spread and median of the data values for different health responses in NY state for the years 2006 and 2010. Comparing the distributions of responses across two different years within the same state, we could see there are some improvement in health perceptions for "Poor", "Fair", "Good" response from 2006 to 2010, where the central line within each box has improved, which represents the median, the middle value of the data. From the interquartile range (the main body of the box plot), we could see there was a greater variability in perceptions with "Poor", "Fair", "Very good" responses, where the boxes are wider.** 

# Problem 3

Import datasets.

```{r}
nhanes_covar =
  # import dataset
  read_csv("data/nhanes_covar.csv", skip = 4) |>
  
  # clean names
  janitor::clean_names() |>
  
  # ensure variables are appropriate encoded
  
  mutate(
   sex = 
      case_match(
        sex,
        1 ~ "Male",
        2 ~ "Female"),
    
    education =
      case_match(
        education,
        1 ~ "Less than high school",
        2 ~ "High school equivalent",
        3 ~ "More than high school")) |>
  
  # exclude participants less than 21 years of age, and those with missing demographic data; 
  filter(age >= 21) |>
  drop_na()
```

 

```{r}
nhanes_accel =
  read_csv("data/nhanes_accel.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    names_prefix = "min",
    values_to = "activity") |>
  mutate(minute = as.numeric(minute))
```


Merge datasets.

```{r}
merged_df = nhanes_covar |>
  inner_join(nhanes_accel, by = "seqn")
```

Produce a reader-friendly table for the number of men and women in each education category, 

```{r}
education_table = merged_df |>
  group_by(sex, education) |>
  summarize(count = n()) |>
  knitr::kable(digits = 1)

print(education_table)
```

Create a visualization of the age distributions for men and women in each education category. 

```{r}
ggplot(merged_df, aes(x = age, fill = sex)) +
  geom_histogram(position = "dodge", binwidth = 2) +
  facet_grid(. ~ education) +
  labs(title = "Age distribution by Sex and Education Level",
       x = "Age",
       y = "Count",
       caption = "Data from the nhanes_covar and nhanes_accel datasets")
```

**Comments: There are more males than females have education of level of high school equivalent. Both males and females have more people with education level of more than high school. The age distribution among high school equivalent and less than high school is approximately same. The age distribution for more than high school is tends to lean on lower ages.**




Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences.

```{r}
activity_df = merged_df |>
  group_by(seqn, sex, age, education) |>
  summarize(total_activity = sum(activity))

ggplot(activity_df, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(. ~ education) +
  labs(title = "Total Activity vs. Age by Education Level",
       x = "Age",
       y = "Total Activity",
       caption = "Data from the nhanes_covar and nhanes_accel datasets") 
```

**Comments: From the plot, we can see the spread of the points mostly stay in the bottom of the plot, which means most of people have lower activity. From the slopes of the trend lines, we could see the negative slopes, which indicates the negative relationship between the two variables of age and total activities, when the age goes up, the total activities goes down, and vice versa. From the plot of high school equivalent and less than high school, we can see males tend to have higher activity. For the more than high school plot, female has a slightly higher trend line than male, which indicates female has a little higher total activity than male. There are also some anomalies points in the plots, which means there are a few people don't follow the trend line and have higher activity.**



Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. 

```{r}
inspection_activity = merged_df 

ggplot(inspection_activity, aes(x = minute, y = activity, color = sex)) +
  geom_line(aes(group = seqn), alpha = 0.2) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education) +
  labs(title = "24-hour Activity Time Course by Education Level",
       x = "Time of Day",
       y = "Activity",
       color = "Sex",
       caption = "Data from the nhanes_covar and nhanes_accel datasets") 
```

Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

**The faint lines are closely clustered, which indicates that many people have similar activity patterns. Except for the plot of more than high school, there are some lines are scattered, which indicates that some people have different activity patterns. The color represent sex, which indicates males and females mostly have similar trends of activity around the day. Except for more than high school, females tend to might be more active during the middle of the day. The smoothed line gives an average trend across everyone. The peaks in those trend suggest that most people have active times during 500 min to 1000 min in the 24-hour, while troughs suggest more sedentary periods. Since the x-axis represents time, activity during late night hours (time0-time250) are likely due to sleeping, peak could be associated with typical working hours, meal times, or exercise periods (time500-time1000+).**